An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging

### 2018-02-09 HowtoHowtoCapon/document.tex
Getting the hang of including articles

%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% CONDITIONALS %
%%%%%%%%%%%%%%%%                                ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\newif\ifPeerReview\PeerReviewtrue              % Whether to create the PeerReview version or
                                                % Journal version
\newif\ifOnlineColor\OnlineColortrue            % Compile online color version?

\newif\ifFlatArchive\FlatArchivefalse           % Whether archive is flat (messy) or contain 
                                                % subfolders for graphics etc.
\newif\ifFloatAtEnd\FloatAtEndfalse             % Available in PeerReview mode:
                                                % Place floats at end of document?
\newif\ifTODO\TODOtrue                        % Use todo notes?

\ifOnlineColor
   \newcommand\figPostfix{_online}              % Colored images online
\else
   \newcommand\figPostfix{_bw}                  % Black and white in journal
\fi

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% IEEEtran %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\ifPeerReview
\documentclass[12pt,journal,draftclsnofoot,onecolumn]{IEEEtran}
% \newcommand\CLASSINPUTbaselinestretch{1.66}     % http://theoval.cmp.uea.ac.uk/~nlct/latex/thesis/node17.html
\else
\documentclass[journal]{IEEEtran}
\fi

% \RequirePackage[latin1]{inputenc}%              % Set input encoding (optionally latin1)
% \RequirePackage[T1]{fontenc}%                 % Set font encoding
% \usepackage[norsk]{babel}

% Single column review mode:
%\documentclass[12pt,journal,onecolumn]{IEEEtran}
%\newcommand\CLASSINPUTbaselinestretch{2}
%\RequirePackage{calc}
%\RequirePackage{fp} 
% Left margin = 1 inch + hoffset + oddsidemargin (or evensidemargin)
% Adding the 2cm gutter width to the odd/even side margins:
%\setlength\hoffset{0pt}
%\setlength\oddsidemargin{4cm}       % 4cm margins on the left side
%\addtolength\oddsidemargin{-1in}    % Subtract the initial 1 inch
%\setlength\textwidth{21cm-8cm}    % A4 width (21cm) minus margins on either side

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% IEEE ''APPROVED'' PACKAGES %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                  ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

\usepackage{cite}

\ifCLASSINFOpdf
   \usepackage[dvips]{graphicx}                 % Might not work. Use 'latex' instead of 
   \ifFlatArchive\else                          % 'pdflatex'
      \graphicspath{./gfx/}
   \fi
\else
   \usepackage[dvips]{graphicx}
   \ifFlatArchive\else
      \graphicspath{./gfx/}
   \fi
\fi

\RequirePackage[table,dvipsnames,svgnames]{xcolor}

\usepackage[cmex10]{amsmath}                    % cmex10 option to be IEEE explore compliant
\interdisplaylinepenalty=2500                   % Allows multiline equations to be broken

% \RequirePackage{amssymb}

\RequirePackage{array}

\ifCLASSOPTIONcompsoc
   \usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
   \usepackage[caption=false,font=footnotesize]{subfig}
\fi
\ifCLASSOPTIONcaptionsoff                       % IEEE promoted hack to turn off captions from the 
   \let\MYorigsubfloat\subfloat                 % subfloat package should the captionsoff option
   \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}} % be specified.
\fi

\ifFloatAtEnd
\ifCLASSOPTIONcaptionsoff                       % Places float at the end of the document when the
  \usepackage[nomarkers]{endfloat}              % captionsoff options is specified to IEEEtrans.cls
  \let\MYoriglatexcaption\caption               % (PeerReview mode)
  \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
\fi
\fi

\usepackage{fixltx2e}                           % Fix some twocolumn float problems

%\usepackage{stfloats}                          % Allows: \begin{figure*}[!b]
                                                % (double column figures on top/bottom)

\usepackage{url}                                % Support for handling and breaking URLs

% NOTE: PDF hyperlink and bookmark features are not required in IEEE
%       papers and their use requires extra complexity and work.
\newcommand\MYhyperrefoptions{bookmarks=true,bookmarksnumbered=true,
pdfpagemode={UseOutlines},plainpages=false,pdfpagelabels=true,
colorlinks=true,linkcolor={black},citecolor={black},urlcolor={black},
pdftitle={An Optimized GPU Implementation of the MVDR Beamformer for Active Sonar Imaging},
pdfsubject={},
pdfauthor={Jo Inge Buskenes},
pdfkeywords={adaptive beamforming, beamforming, complexity, sonar, active}}%
\ifCLASSINFOpdf
   \usepackage[\MYhyperrefoptions,pdftex]{hyperref}
\else
   \usepackage[\MYhyperrefoptions,breaklinks=true,dvips]{hyperref}
   \usepackage{breakurl}                        % Allows 'dvips' driver to break links
\fi

%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% ADDITIONAL PACKAGES %       
%%%%%%%%%%%%%%%%%%%%%%%                         ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \ifPeerReview
% \let\OldIncludegraphics{\includegraphics}
% \usepackage{letltxmacro}
% \LetLtxMacro{\OldIncludegrsaphics}{\includegraphics}
% \renewcommand{\includegraphics}[2][]{\OldIncludegraphics[width=\linewidth, #1]{#2}}
% \fi

% % \makeatletter
% % \def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
% % \else\Gin@nat@width\fi}
% % \makeatother
% % \let\Oldincludegraphics\includegraphics
% % \renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}

\usepackage[maxfloats=40]{morefloats}
\newcounter{todoidx}
% \setcounter{todoidx}

\ifTODO
   \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
   \setlength\marginparsep{1pt}
   \setlength\marginparwidth{35pt}
   \newlength\marginparwidthsmall
   \setlength\marginparwidthsmall{\marginparwidth}
   \addtolength\marginparwidthsmall{-7pt}
   \newcommand\todo[1]{%
      \addtocounter{todoidx}{1}%
      {\color{Red}\bf(\thetodoidx{})}%%\fbox{\bf\thetodoidx{}}}%
      \marginpar{%
         {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
         \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{\scriptsize #1}}}}

   \newcommand\todopar[1]{\fcolorbox{red}{white}{\parbox{0.97\linewidth}{#1}}}
\else
%    \usepackage[disable]{./todonotes} 
   \newcommand\todo[1]{}
\fi

\newenvironment{narrow}[2]{%
\begin{list}{}{%
\setlength{\topsep}{0pt}%
\setlength{\leftmargin}{#1}%
\setlength{\rightmargin}{#2}%
\setlength{\listparindent}{\parindent}%
\setlength{\itemindent}{\parindent}%
\setlength{\parsep}{\parskip}}%
\item[]}{\end{list}}

\usepackage{float}

\ifOnlineColor
   \definecolor{tabBlue}{HTML}{AACCFF}
\else
   \definecolor{tabBlue}{HTML}{CCCCCC}
\fi

%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% MACROS %       
%%%%%%%%%%                                      ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \DeclareMathOperator*{\argmin}{\text{arg}\;\text{min}}

\newcommand\Fig[1]{Fig.~\ref{#1}}

\newcommand\Grey[1]{{\color{Grey}#1}}
\newcommand\Red[1]{{\color{Red}#1}}
\newcommand\Blue[1]{{\color{Blue}#1}}
\newcommand\DarkBlue[1]{{\color{DarkBlue}#1}}
\newcommand\LightBlue[1]{{\color{LightBlue}#1}}
\newcommand\Brown[1]{{\color{Brown}#1}}
\newcommand\Green[1]{{\color{Green}#1}}
\newcommand\SeaGreen[1]{{\color{SeaGreen}#1}}
\newcommand\Yellow[1]{{\color{yellow}#1}}
\newcommand\Orange[1]{{\color{orange}#1}}

\newcommand\nn{\nonumber\\}

\newcommand\nmat[1]{\begin{matrix}#1\end{matrix}}
\newcommand\bmat[1]{\begin{bmatrix}#1\end{bmatrix}}
\newcommand\case[1]{\begin{cases}#1\end{cases}}
\newcommand\textbox[2]{\footnotesize\text{\parbox{#1}{\centering\emph{#2}}}}

\newcommand\rand{\text{rand}}
\newcommand\randn{\text{randn}}
\newcommand\rect{\text{rect}}
\newcommand\sinc{\text{sinc}}
\newcommand\tr{\text{tr}}
\newcommand\adj{\text{adj}}

% \newcommand\max{\text{max}}
\newcommand\argmin[1]{\text{arg}\;\underset{#1}{\text{min}}}

\newcommand\qqquad{\quad\qquad}
\newcommand\qqqquad{\qquad\qquad}

% \renewcommand\l[1]{\left#1}
% \renewcommand\r[1]{\right#1}

% {\text{\parbox{1.5cm}{\centering volume hyper- sphere}}}

%Keyword colouring:
\newcommand\kw[1]{#1}
\newcommand\parm[1]{#1}%\color{Black}#1\color{Black}}

\newcommand\of[1]{\scriptstyle(\parm{#1})\displaystyle}
\newcommand\df[1]{\scriptstyle[\parm{#1}]\displaystyle}
\newcommand\var[3]{#1_\text{#2}\of{#3}}

\newcommand\diag{\text{diag}}

% \raisebox{lift}[extend-above-baseline][extend-below-baseline]{text}
\newcommand\mt[1]{\text{\emph{#1}}} %mt = mathtext
\newcommand\mathnorm{\textstyle}
\newcommand\mathbig[1]{\displaystyle#1\mathnorm}
\newcommand\mathsmall[1]{\scriptstyle#1\mathnorm}
\newcommand\mathtiny[1]{\scriptscriptstyle#1\mathnorm}
\newcommand\sfrac[2]{\scriptstyle\raisebox{0.25pt}[0pt][0pt]{$\frac{#1}{#2}$}\mathnorm}
\newcommand\nfrac[2]{\textstyle\frac{#1}{#2}\displaystyle}

\newcommand\sumu[1]{\sum\limits^{#1}\;}
\newcommand\suml[1]{\sum\limits_{#1}\;}
\newcommand\sumb[2]{\sum\limits_{#1}^{#2}\;}

\newcommand\produ[1]{\prod\limits^{#1}\;}
\newcommand\prodl[1]{\prod\limits_{#1}\;}
\newcommand\prodb[2]{\prod\limits_{#1}^{#2}\;}

\newcommand\defeq{\overset{\underset{\mathrm{def}}{}}{=}}

%Math macros:
\newcommand\T{^{\scriptscriptstyle T}}
\renewcommand\H{^{\scriptscriptstyle H}}

\renewcommand\vec[1]{\boldsymbol{#1}}
\newcommand\mat[1]{\boldsymbol{#1}}

\newcommand\Om{O_\text{m}}
\newcommand\Oa{O_\text{a}}
\newcommand\Nl{N_\text{l}}
\newcommand\Nk{N_\text{k}}
\newcommand\1{\vec 1}
\newcommand\I{\mat I}
\renewcommand*\a{\vec a}
\renewcommand*\i{\vec i}
\renewcommand*\k{\vec k}
\newcommand*\n{\vec n}
\newcommand*\p{\vec p}
\newcommand*\s{\vec s}
\newcommand*\w{\vec w}
\newcommand*\x{\vec x}
\newcommand*\y{\vec y}

\newcommand*\A{\mat A}
\newcommand*\B{\mat B}
\newcommand*\C{\mat C}
\newcommand*\E{\mat E}
\renewcommand*\P{\mat P}
\newcommand*\eP{\mat{\hat P}}
\newcommand*\R{\mat R}
\newcommand*\Ri{\R^{-1}}
\newcommand*\eR{\mat{\hat R}}
\newcommand*\eRi{\hat{\mat R}\;\!^{-1}}
\newcommand*\Navg{N_\text{avg}}
\newcommand*\W{\mat W}
\newcommand*\X{\mat X}
\newcommand*\Xd{\X_{\!\Delta}}
\newcommand*\Y{\mat Y}

\renewcommand*\L{\mat \Lambda}
\newcommand*\U{\mat U}
% \renewcommand*\t{\mathtiny{^T}}
% \newcommand*\h{\mathtiny{^H}}
\renewcommand*\t{^T}
\newcommand*\h{^H}

\usepackage{tikz}
\usetikzlibrary{shapes,snakes}
\usepackage{amsmath,amssymb}
% \usepackage{datatool}
% \usepackage{glossaries}

\newenvironment{outline}
{\begin{itemize}}
{\end{itemize}}

%    \definecolor{todobackground}{rgb}{0.95,0.95,0.95}
%    \setlength\marginparsep{3pt}
%    \setlength\marginparwidth{42pt}
%    \newlength\marginparwidthsmall
%    \setlength\marginparwidthsmall{\marginparwidth}
%    \addtolength\marginparwidthsmall{-7pt}
%    \newcommand\todo[1]{%
%       \addtocounter{todoidx}{1}%
%       {\color{Red}\fbox{\bf\thetodoidx{}}}%
%       \marginpar{%
%          {\vspace*{-10pt}\color{Red}\fbox{\bf\thetodoidx{}}}\\%
%          \fcolorbox{red}{todobackground}{\parbox{\marginparwidthsmall}{#1}}}}
%

% correct bad hyphenation here
% \hyphenation{op-tical net-works semi-conduc-tor}

%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% GLOSSARY %
%%%%%%%%%%%%                                    ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \makeglossaries
% \input{../../Library/glossary}

%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% DOCUMENT START %
%%%%%%%%%%%%%%%%%%                              ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \usepackage{yfonts}

\graphicspath{submission/final/gfx/}

\begin{document}

@@

\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}sen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

@@

The Minimum Variance Distortionless Response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR on a Graphics Processing Unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s). Compared to our initial CPU implementation the optimizations described herein led to a speedup of more than 2 orders of magnitude, or an expected 5-10 times improvement had the CPU received similar optimization effort. This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
% If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.

@@

% Keywords (normally not used for peer reviews)
\ifPeerReview\else
\begin{IEEEkeywords}
Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
\end{IEEEkeywords}
\fi}
% \fi

@@

\IEEEPARstart{D}{ata} driven methods have been introduced in various imaging fields over the years in attempts to improve image quality. One such method is the Minimum Variance Distortionless Response (MVDR) beamformer. When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in e.g. radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2012a,Blomberg2011,Dursun2009,Lo2004}.

@@

% Robustification:
% R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% -  
To ensure MVDR robustness, the literature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}. The complexity issue, on the other hand, can be handled by introducing well-founded approximations. For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}. Other studies perform MVDR beamforming in beamspace (the spatial frequency domain), which can be considered sparse due to the limited angular extent of the received beam~\cite{VanTrees2002,Nilsen2009}. Exploiting this can lead to considerable performance improvements, especially in high channel count systems.

@@

In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by 1-2 orders of magnitude by implementing MVDR on a Graphics Processing Unit (GPU). Similar work may be found in e.g. medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}. While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. Our implementation is not restricted in this manner.

Our implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field. The system has a range resolution comparable to the sample period so our MVDR implementation runs in near single snapshot mode. To deal with this and to avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}. This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make it feasible to use MVDR in an active system. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive methods~\cite{Buskenes2013}.

This article is outlined as follows: In section \ref{methods} we introduce the concept of adaptive beamforming and provide details on the MVDR method. Then in \ref{maptogpu} we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in \ref{images_and_benchmarks}-\ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

@@

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then - before summation - a suitable set of delays, $\Delta$, and weights, $w$, are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

By definition the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\} \in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985},
\item \emph{temporal averaging} over an interval comparable to the pulse length (1-5 samples) to maintain true speckle statistics~\cite{Synnevag2009a}, and
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$,
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as:
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

@@

@@

Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require 3 times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

@@

\bibitem{Blomberg2011}
A.~E.~A. Blomberg, R.~E. Hansen, S.~A.~V. Synnes, and A.~Austeng, ``{Improved
  interferometric sonar performance in shallow water using adaptive
  beamforming},'' in \emph{Proceedings of the International Conference \&
  Exhibition on Underwater Acoustic Measurements (UAM), Kos, Greece, June},
  2011.

@@

\end{document}

>>>>

@@

\author{Jo~Inge~Buskenes,~\IEEEmembership{Student Member,~IEEE,} %
        Jon~Petter~\AA{}Åsen,~\IEEEmembership{Student Member,~IEEE,} %
        Carl-Inge~Colombo~Nilsen,~\IEEEmembership{Member,~IEEE,} %
        Andreas~Austeng,~\IEEEmembership{Member,~IEEE}%
\thanks{J. I. Buskenes, C.-I. C. Nilsen and A. Austeng are with the Department of Informatics, University of Oslo, 0316 Oslo, Norway.}%
\thanks{J. P. \AA{}sen is with MI-Lab, Norwegian University of Science and Technology, 7491 Trondheim, Norway.}% <-this % stops a space
% \thanks{Manuscript received April 19, 2005; revised January 11, 2007.}
}

@@

The minimum variance distortionless response (MVDR) beamformer has recently been proposed as an attractive alternative to conventional beamformers in active sonar imaging. Unfortunately, it is very computationally complex because a spatial covariance matrix must be estimated and inverted for each image pixel. This may discourage its use unnecessarily in sonar systems which are continuously being pushed to ever higher imaging ranges and resolutions.

In this study we show that for active sonar systems up to 32 channels, the computation time can be significantly reduced by performing arithmetic optimizations, and by implementing the MVDR beamformer on a graphics processing unit (GPU). We point out important hardware limitations for these devices, and assess the design in terms of how efficiently it is able to use the GPU's resources. On a quad-core Intel Xeon system with a high-end Nvidia GPU, our GPU implementation renders more than a million pixels per second (1\;MP/s). Compared to our initial central processing unit (CPU) implementation the optimizations described herein led to a speedup of more than two orders of magnitude, or an expected five to ten times improvement had the CPU received similar optimization effort. This throughput enables real-time processing of sonar data, and makes the MVDR a viable alternative to conventional methods in practical systems.
% If the HISAS1030 was attached to a platform traveling at 1.8m/s, creating full-coverage sectorscan images in real-time would be possible with this performance.

@@

% Keywords (normally not used for peer reviews)
\begin{IEEEkeywords}
Adaptive beamforming, MVDR, Capon, sonar, active sonar, GPU, graphics processing unit, complexity.
\end{IEEEkeywords}
}

@@

\IEEEPARstart{D}{ata-driven} methods have been introduced in various imaging fields over the years in attempts to improve image quality. One such method is the minimum variance distortionless response (MVDR) beamformer. When compared to conventional methods, MVDR is often able to improve image contrast and resolution, as shown in, for example, radar~\cite{Benitz1997}, ultrasound~\cite{Synnevag2007}, and active sonar~\cite{Blomberg2013,Blomberg2012a,Dursun2009,Lo2004}.

@@

% Robustification:
% R estimate accuracy vs. dynamic trackability: (1) Trade observation time for bandwidth (2) Partial adaptiveness (less degrees of freedom than channels) (3) Quadratic constraint on sensor weights to limit the norm to some preset value.
% -  
To ensure MVDR robustness, the literature suggests combining means such as temporal and spatial averaging, and regularization~\cite{VanTrees2002,Kailath1985}. On the other hand, the complexity issue can be handled by introducing well-founded approximations. For instance, some studies assume data stationarity which allows the formation of a Toeplitz-structured covariance matrix that is simpler to invert~\cite{Asl2012a,Jakobsson2000}. Other studies perform MVDR beamforming in beamspace (the spatial frequency domain), which can be considered sparse due to the limited angular extent of the received beam~\cite{VanTrees2002,Nilsen2009}. Exploiting this can lead to considerable performance improvements, especially in high channel count systems.

@@

In this work, we show that such approximations may not be necessary for sonar systems up to 32 channels, since here the computation time can be reduced by one to two orders of magnitude by implementing MVDR on a graphics processing unit (GPU). Similar work may be found in, for example, medical ultrasound imaging, where Chen \emph{et al.} have demonstrated a GPU implementation of the MVDR that operates on real valued data~\cite{Chen2011,Chen2011a}. While this implementation is fast, the real valued data constraint does not allow the creation of asymmetric and shifted responses, and hence prevents the MVDR's potential to be fully reached. Our implementation is not restricted in this manner.

Our implementation is adapted to a well sampled 100\;kHz wideband active sonar system operating in the near field. The system has a range resolution comparable to the sampling period so our MVDR implementation runs in near single snapshot mode. To deal with this and to avoid signal cancellation we compute and average the covariance estimates from overlapping subarrays~\cite{Kailath1985}. This may resemble techniques such as "subarray MVDR"~\cite{Chapman1976}, but unlike that technique we are not using subarrays to reduce complexity but to make it feasible to use MVDR in an active system. In related work, we investigated using the same MVDR GPU implementation in cardiac ultrasound imaging~\cite{Asen2012,Asen2013}, and in active sonar imaging we have compared the performance of this implementation to comparable adaptive beamformers~\cite{Buskenes2013}.

This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR method. Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

@@

\ifPeerReview
\begin{figure}[H]\centering
\includegraphics[width=0.8\linewidth]{gfx/buske1\figPostfix.eps}
\else
\begin{figure}[!t]\centering
\includegraphics[width=\linewidth]{gfx/beamforming.eps}
\fi%
\caption{Beamforming principle. Signal signature is first removed by matched filtering. Then, before summation, a suitable set of delays $\Delta$ and weights $w$ are applied to focus on a pixel of interest at angle and range ($\theta,n$).}\label{beamforming}
\end{figure}
Let the receiver be an $M$ element uniform linear array, and assume that the signature of the transmitted signal has been removed by a matched filter (Fig. \ref{beamforming}). Further assume that the array channels are digitally delayed to focus at a pixel with azimuth angle $\theta$ and range sample $n$, such that the delayed data from the $m$th channel can be expressed as $x_m[\theta,n]$. To simplify notation, we make the dependence on $\theta$ implicit from now on.

By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

We perform \emph{temporal} and \emph{spatial averaging} first and put the result in an intermediate sample covariance matrix $\breve{\R}$. To do this we need to segment our array into subarrays. If we let $x_l[n]$ represent the data vector from subarray $l$ by
\begin{gather}
\x_l[n] = \bmat{x_l[n] & x_{l+1}[n] & \dots & x_{l+L-1}[n]}\T,
\end{gather}
then $\breve{\R}$ can be calculated as
\begin{gather}
\breve{\R}[n] =  \frac{1}{N_K N_L} \sumb{l=0}{M-L}\sumb{n'=n-K}{n+K} \x_l[n']\x_l\H[n'] \in\mathbb{C}^{L,L},\label{spatialR}
\end{gather}
where $N_K = 2K+1$ is the number of temporal samples to perform averaging over, and $N_L = M-L+1$ is the number of subarrays.

@@

% \the\linewidth

@@

Note how the computation of $\eR$ completely dominates at smaller subarray sizes, that solving $\eRi\1$ only plays a notable role for larger array and subarray sizes, and that the computation of $z$ has a negligible impact on computation time. Also notice how temporal averaging comes with a high computational penalty. This is because building $\eR$ is heavy on complex multiplications, which require three times as many arithmetic instructions as a complex addition, and a lot of these are repeated unnecessarily.

@@

@@
